Declarative deployment reconfiguration of pulverised collective adaptive systems at runtime \thanks{Identify applicable funding agency here. If none, delete this.}

### 2023-05-16 paper-2023-acsos-pulverization.tex
write something before evacuation

In principle, there are several possibilities to implement a pulverisation configurator and runtime:
as a library
(configured and executed by a target programming language),
as a framework/runtime (configured by means of configuration files and with its own runtime),
or as a \ac{dsl} (configured with its own custom syntax).
%
The approaches have different trade-offs.
%
A (well designed) library can be flexible and easier to adopt
as it does not require to learn new languages;
but it will be usable only in the language/runtime it has been designed for,
and will require its own framework/runtime to be executed
(or appropriate bindings for existing ones).
%
A framework/runtime configured via declarative files enforces declarativity at the configuration level,
but it usually dictates network protocols and features limited flexibility,
as options that have not been accounted for at design time can hardly be injected through configuration files.
%
Finally, a \ac{dsl} can be designed to be as expressive as needed,
but has high maintenance cost and, potentially, a steep learning curve due to the need of learning a new (custom) language.
%

TODO: recently, internal DSLs

>>>>

In principle, there are several possibilities to implement a pulverisation configurator and runtime:
as a library
configured and executed by a target programming language,
as a framework/runtime configured by means of configuration files and with its own runtime,
or as a \ac{dsl} configured with its own custom syntax.
%
These approaches have different trade-offs.
%
A (well designed) library can be flexible and easier to adopt
as it does not require to learn new languages;
but it will be usable only in the language/runtime it has been designed for,
and will require its own framework/runtime to be executed
(or appropriate bindings for existing ones).
%
A framework/runtime configured via declarative files enforces declarativity at the configuration level,
but it usually dictates network protocols and features limited flexibility,
as options that have not been accounted for at design time can hardly be injected through configuration files.
%
Finally, a \ac{dsl} can be designed to be as expressive as needed,
but it has high maintenance cost (as the language maintenance costs stack upon those of the library/API) and, potentially,
a steep learning curve due to the need of learning a new (custom) language.
%
Recent evolution in programming languages,

such as Kotlin, Ruby, and Scala~\cite{Riti2018}
however,
a trend has emerged that
TODO: recently, internal DSLs
adopted by gradle
